---
title: "SEA-AD Prep for 2025 Jax workshop"
author: "Robert R Butler III"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

To prep data for the workshop, it will need to be converted to Seurat & Giotto formats
```{r, load, include=FALSE}
library(Seurat)
library(anndata)
# library(anndataR)
library(data.table)

# In Rstudio both add giotto_env to LD and configure newer GEOS to work
Sys.unsetenv("RETICULATE_PYTHON")
Sys.setenv(
  LD_LIBRARY_PATH = paste0(
    "/labs/flongo/reference/envs/giotto_env/lib:",
    "/scg/apps/software/GEOS/3.11.1/lib64:",
    Sys.getenv("LD_LIBRARY_PATH")
  )
)
dyn.load("/scg/apps/software/GEOS/3.11.1/lib64/libgeos.so.3.11.1")
dyn.load("/scg/apps/software/GEOS/3.11.1/lib64/libgeos_c.so")

reticulate::use_python(
  required = TRUE, 
  python = "/labs/flongo/reference/envs/giotto_env/bin/python"
)
reticulate::use_condaenv(
  condaenv = "/labs/flongo/reference/envs/giotto_env/bin/python",
  conda = "~/miniconda3/condabin/conda"
)


library(Giotto)
library(qs)

```

## Look at the metadata
```{r, get_meta}
# read in metadata 
meta <- data.table(readxl::read_xlsx("Supplementary Table 1.xlsx"))
meta <- meta[`snRNA-seq` == "Y" & MERFISH2 == "Y"]
meta <- meta[is.na(`If other Consensus dx, describe`)]
meta$`Overall AD neuropathological Change` <- factor(
  meta$`Overall AD neuropathological Change`, levels = c("Not AD", "Low", "Intermediate", "High")
)

sc_meta <- fread("SEAAD_MTG_RNAseq_all-nuclei_metadata.2024-02-13.csv")
sc_meta <- unique(sc_meta[, .(`Donor ID`, `Continuous Pseudo-progression Score`)])
meta <- merge(meta, sc_meta, all.x = TRUE)
remove(sc_meta)
gc()

# subsetting for donors on extremes, without comorbidities
view_cols <- c(
  "Donor ID",
  "Age at Death",
  "Sex", 
  "Continuous Pseudo-progression Score",
  "Overall AD neuropathological Change",
  "Thal",
  "Braak",
  "CERAD score",
  "Overall CAA Score",
  "Highest Lewy Body Disease",
  "Cognitive Status",
  "Last MMSE Score",
  "APOE Genotype",
  names(meta)[names(meta) %like% "Consensus Clinical Dx"]
)
setorderv(meta, view_cols[4])
View(meta[, ..view_cols])

donors <- meta[
  (
    `Overall AD neuropathological Change` %in% c("Not AD", "Low") 
    & `Continuous Pseudo-progression Score` < 0.35
    & `Highest Lewy Body Disease` != "Neocortical (Diffuse)"
  )
  | (
    `Overall AD neuropathological Change` %in% c("Intermediate", "High") 
    & `Continuous Pseudo-progression Score` > 0.6
  ),
][order(get(view_cols[4])), `Donor ID`]

spat_donors <- donors[c(3, 6, 5, 10:12)]

```


## Prep the (cleaned) snRNAseq data
This will define standard settings you intend to run for the analysis
```{r, get_sc_data}
# set up the dir and metadata to keep
setwd("/labs/flongo/reference/SEA-AD")
cols_keep <- c(
  "Donor ID",
  "Neurotypical reference",
  "Sex",
  "Age at Death",
  "Overall AD neuropathological Change",
  "Thal",
  "Braak",
  "CERAD score",
  "Overall CAA Score",
  "Highest Lewy Body Disease",
  "Cognitive Status",
  "Last MMSE Score",
  "Interval from last MMSE in months",
  "Continuous Pseudo-progression Score",
  "APOE Genotype",
  "Class",
  "Subclass",
  "Supertype"
)

# run a loop for all samples
sc_list <- lapply(donors, function(i) {
  # from the aws host
  fn <- paste0(i, "_SEAAD_MTG_RNAseq_final-nuclei.2024-02-13.h5ad")
  print(sprintf("downloading %s...", fn))
  download.file(
    url = paste0(
      "https://sea-ad-single-cell-profiling.s3.amazonaws.com/MTG/RNAseq/donor_objects/",
      fn
    ),
    destfile = fn,
    quiet = TRUE,
    method = "wget", 
    extra = "-N"
  )
  
  # build the Seurat Object with subset of the metadata
  sc <- read_h5ad(fn)
  sc1 <- CreateSeuratObject(
    counts = t(sc$X),
    assay = "RNA",
    meta.data = sc$obs[, cols_keep],
  )
  
  # downsample to save space (5k cells per sample)
  set.seed(123)
  cells <- sample(Cells(sc1), size = 2000)
  sc1 <- subset(sc1, cells = cells)
  
  # return and manage memory
  remove(sc)
  gc()
  return(sc1)
})


# Join the samples
sc <- merge(
  x = sc_list[[1]],
  y = sc_list[2:length(sc_list)]
) 
remove(sc_list)
gc()
sc <- JoinLayers(sc)

# Add a Group column
Idents(sc) <- "Donor.ID"
sc[["Group"]] <- as.character(
  plyr::mapvalues(
    x = sc@active.ident,
    from = donors,
    to = c(rep('ctrl', 6), rep('ad', 6))
  )
)

# Save output
saveRDS(sc, file = paste("SEAAD_MTG_RNAseq_final-nuclei.2k_demo.rds", sep = "."))
remove(sc)
gc()

```

## Prep the (cleaned MERFISH data)
```{r}
# turns out there is some issue with the way rstudio uses reticulate that this 
# will only work inside R, because who the hell knows why  https://github.com/dynverse/anndata/issues/15 
# i don't really want to play around with reticulate and break it just to get it to maybe work for rstudio

# read in
cols_keep <- cols_keep[!cols_keep %in% c("Neurotypical reference", "APOE Genotype")]
sc <- read_h5ad("SEAAD_MTG_MERFISH.2024-12-11.h5ad")
# i <- spat_donors[1]
# a <- data.table(sc$X[sc$obs[["Donor ID"]] == i,], keep.rownames = TRUE)
# b <- data.table(
#   sc$obs[sc$obs[["Donor ID"]] == i, c("Cell ID", cols_keep)], 
#   keep.rownames = "cell_ID"
# )
# # checked and `spatial` is the same as `X_spatial_raw`
# c <- data.table(sc$obsm[['spatial']][sc$obs[["Donor ID"]] == i,])

# choose one section per sample
set.seed(456)
sects <- unique(
  data.table(
    sc$obs[
      sc$obs[["Donor ID"]] %in% spat_donors & !is.na(sc$obs[["Merscope"]]), 
      c("Donor ID", "Merscope", "Section")
    ]
  )
)
sects <- sects[sects[ , .I[sample(.N, 1)], by = "Donor ID"]$V1, Section]

# set_giotto_python_path(python_path = python_path, verbose = TRUE)
# by directly saving plots, but not rendering them you will save a lot of time
instrs <- createGiottoInstructions(
  save_plot = TRUE,
  show_plot = FALSE, 
  return_plot = FALSE
)

# load each sample into a separate giotto object
gobjs <- lapply(spat_donors, function(i) {
  # Create Giotto object
  createGiottoObject(
    expression = t(
      sc$X[
        sc$obs[["Donor ID"]] == i 
        & sc$obs[["Section"]] %in% sects 
        & !is.na(sc$obs[["Merscope"]]),
      ]
    ), 
    spatial_locs = sc$obsm[['X_spatial_raw']][
      sc$obs[["Donor ID"]] == i 
      & sc$obs[["Section"]] %in% sects
      & !is.na(sc$obs[["Merscope"]]),
    ],
    cell_metadata = data.table(
      sc$obs[
        sc$obs[["Donor ID"]] == i 
        & sc$obs[["Section"]] %in% sects
        & !is.na(sc$obs[["Merscope"]]),
        cols_keep
      ],
      keep.rownames = "cell_ID"
    ),
    instructions = instrs
  )
})

# cleanup anndata (may need to restart R session if lagging too badly)
remove(sc)
gc()
saveRDS(gobjs, file = "temp_gobjs.rds")
# gobjs <- readRDS("temp_gobjs.rds")

# rotate each to the same orientation
# spatPlot(spin(gobjs[[6]], 0), cell_color = "Subclass", show_plot = TRUE)
spins <- c(-85, -71, -13, -44, 6, 94) 
gobjs <- lapply(seq(gobjs), function(i) {
  spin(gobjs[[i]], angle = spins[i])
})

# shift each slide to the bottom-left most global coords (0,0) prior to merging
gobjs <- lapply(gobjs, function(i) {
  coords <- ext(i)
  spatShift(i, dx = -coords$xmin, dy = -coords$ymin)
})

# merge the slides together -----------------------------
# lapply(gobjs, getSpatialLocations)
gobj <- joinGiottoObjects(
  gobjs,
  gobject_names = spat_donors,
  join_method = "shift",
  x_shift = c(0, 3700, 5900, 9200, 11700, 14800),
  y_shift = NULL,
  x_padding = NULL,
  y_padding = NULL,
  verbose = TRUE
)

# drop extra levels of each factor
gobj@cell_metadata$cell$rna$`Donor ID` <- factor(gobj@cell_metadata$cell$rna$`Donor ID`)

# Add a Group column
meta <- pDataDT(gobj)[, .(cell_ID, `Donor ID`)]
meta <- merge(
  x = meta,
  y = data.table("Donor ID" = donors, Group = c(rep('ctrl', 6), rep('ad', 6))),
  sort = FALSE
)
gobj <- addCellMetadata(gobj, new_metadata = meta[, .(cell_ID, Group)])

# plot to see
spatPlot(gobj,
  cell_color = "Subclass",
  point_size = 1,
  cell_color_code = as.vector(palette.colors(n = 24, palette = "Alphabet")),
  save_param = list(
    save_name = paste("SEAAD_MTG_MERFISH", "spatPlot2D", sep = ".")
  )
)

# save to giotto folder output
saveGiotto(gobj,
  foldername = 'SEAAD_MTG_MERFISH.2024-12-11.giotto_obj',
  method = 'qs',
  overwrite = FALSE
)


# # just the separate groups
# ad <- joinGiottoObjects(
#   gobjs[4:6],
#   gobject_names = spat_donors[4:6],
#   join_method = "shift",
#   x_shift = c(0, 3400, 6100),
#   y_shift = NULL,
#   x_padding = NULL,
#   y_padding = NULL,
#   verbose = TRUE
# )
# ctrl <- joinGiottoObjects(
#   gobjs[1:3],
#   gobject_names = spat_donors[1:3],
#   join_method = "shift",
#   x_shift = c(0, 2300, 5600),
#   y_shift = NULL,
#   x_padding = NULL,
#   y_padding = NULL,
#   verbose = TRUE
# )
# 
# # drop extra levels of each factor
# ad@cell_metadata$cell$rna$`Donor ID` <- factor(ad@cell_metadata$cell$rna$`Donor ID`) 
# ctrl@cell_metadata$cell$rna$`Donor ID` <- factor(ctrl@cell_metadata$cell$rna$`Donor ID`) 
# 
# # plot to see
# lapply(c("ad", "ctrl"), function(i){
#   spatPlot(get(i), 
#     cell_color = "Subclass", 
#     point_size = 1,
#     cell_color_code = as.vector(palette.colors(n = 24, palette = "Alphabet")),
#     save_param = list(
#       save_name = paste(i, "spatPlot2D", sep = ".")
#     )
#   )
# })

# # save to giotto folder output
# saveGiotto(ad, 
#   foldername = 'SEAAD_MTG_MERFISH.2024-12-11.ad.giotto_obj', 
#   method = 'qs', 
#   overwrite = FALSE
# )
# saveGiotto(ctrl, 
#   foldername = 'SEAAD_MTG_MERFISH.2024-12-11.ctrl.giotto_obj', 
#   method = 'qs', 
#   overwrite = FALSE
# )


```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
